Spatial distribution of environmental DNA in a nearshore marine habitat

### 2017-01-13 Manuscript/v1.1_PeerJ/response/response_to_reviews.tex

\maketitle

% \noindent
% {
\begin{tabular}{ l l }
Manuscript Title: &
Spatial distribution of environmental DNA in a nearshore marine habitat
\\
Journal: &
PeerJ
\\
Manuscript ID: &
14649
\\
Authors: &
O'Donnell JL,
Kelly RP,
Shelton AO,
Samhouri JF,
Lowell NC,
Williams GD
\\
\end{tabular}

\bigskip
% }

We are grateful to the editor and reviewers for their insightful feedback.
Reviewer feedback is below in regular typeface, with our responses indented and colored.
\textbf{Line numbers refer to the track changes version of the manuscript}.
% Additional changes to the manuscript independent of the reviews are presented at the end of this document.

@@

%------------------------------------EDITOR-------------------------------------
\subsection*{Editor}
Both reviewers commented favourably on this manuscript and have made some useful and constructive comments which you should consider and respond to. I would also like you to respond to the comment by Bernd Haenfling on your pre-print as he is in agreement with reviewer 1 in relation to the degree of novelty this work represents (line 100 and your concluding sentence). Personally, I found your discussion of the results in the light of the advantages and limitations of this methodology interesting and highly informative. I wondered what the community analyses results would look like if you compartmentalised by e.g. life history or type (e.g. fish)?

\begin{response}
  Thank you for the positive feedback.
	We thank the commenters for pointing out recent literature that we’d missed when developing the original manuscript.
	We have now altered the language in our paper to avoid suggesting that our study is the first to do any kind of explicit spatial analysis, but (as noted below in the response to the commenter), because we have reservations about the way the referenced study was conducted, we have avoided citing it as an example.
	However, the suggestions of reviewer 1 (Wangensteen) on this matter are well taken and we have added the appropriate citations and corrected our wording.
	\\

	Like you, we were curious about the effect of the data subset you mention, and conducted that analysis using the life history data. Unfortunately, we found no meaningful patterns, and have made an attempt to clarify this in the text (L356).
	\\
\end{response}

@@

\subsubsection*{Experimental design}
The objectives are clearly established and the experimental design is perfect for addressing the proposed questions. The laboratory methods are rigorous and the analysis pipeline includes steps for assuring excellent quality of the resulting data set.
\begin{response}
	Thank you; we agree.\\
\end{response}

\subsubsection*{Validity of the findings}
The findings are very robust. The high level of replication in the PCR analyses and the high-standards bioinformatic filtering steps used for removing any sequence attributable to possible contamination assure an excellent resulting data set to work with. The statistical analysis of the data are very clearly explained and the results are highly trustworthy.
\begin{response}
	Thank you; we agree.\\
\end{response}

\subsubsection*{Comments for the Author}
The manuscript is, in general, very well written and the robust results obtained from the enhanced laboratory procedures and bioinformatic pipelines used are presented in a clear manner. I think that the introduction and material and methods sections are a little too long, but they can be enhanced by deleting some sentences (detailed below).
\begin{response}
	Thank you; please see our previous comment on manuscript length, as well as our responses to your specific comments on extraneous details.\\
\end{response}

I think that it would be informative to add a paragraph in the discussion section commenting the caveats of using 16S as metabarcoding marker for taxonomic assignment of the MOTUs (i.e. the issues with the assignment of many sequences which can be assigned only to the family or lower levels), compared to other more variable markers with denser reference databases, such as COI, which, in most cases, allow assignments to the genus or species level.
\begin{response}
	We agree with your point that a locus with only coarse-scale resolution (16S) may obfuscate patterns of diversity compared to a locus with fine-scale resolution. A subset of us has recently published an in-depth comparison of multiple loci for metabarcoding, and we have added a more substantial comment on this and referenced to this now-published work (L452).
\\
\end{response}

@@

\subsubsection*{Specific Corrections}
Some specific corrections are detailed below:

L 36 to 40. This whole paragraph may be deleted to shorten the introduction, without any significant information loss.
\begin{response}
  We can understand the reviewer's perspective, but feel this point is a stylistic one and would prefer to leave the manuscript as it is currently worded to increase the appeal of our work to a broader audience.\\
\end{response}

L90. Correct ``Actiniaria'' and ``Bryozoans''.
\begin{response}
  Thanks for catching these errors. We made these changes (L99, L100).\\
\end{response}

L91. Delete ``(Decapoda)''. These habitats hold many other Crustacea other than Decapoda as well. Actually, I would write ``barnacles (Sessilia) and other crustaceans'' in line 89.
\begin{response}
  Excellent suggestion; we made this change (L98).\\
\end{response}

L93: ``which provide''
\begin{response}
  We made this change (L102).\\
\end{response}

L94: The microscopic plankton is not composed just by ``diatoms and larvae''. Many important holoplanktonic organisms (e.g. copepods) could be mentioned here.
\begin{response}
  We did not mean to imply that the microscopic plankton is composed solely of diatoms and larve, we made a  statement and qualified it with two examples to give meaning and context: `a diverse assemblage of microscopic plankton including diatoms and larvae'.
	We mention diatoms because they are a widely-known, numerically abundant, and ecologically important group that falls outside the scope of this study on metazoans.
	We mention larvae because it subtly hints at our major finding that our samples captured many organisms best known for their sessile, benthic stages (e.g. barnacles).
	We added copepods to this list, though we recognize that it is necessarily incomplete (L104).\\
\end{response}

L100: I am not sure that this is ``the first explicitly spatial analysis of eDNA-derived community similarity'', since some other papers have been published which reported patterns of spatial variability in marine systems (although using less quantitative approaches); e.g. De Vargas et al. 2015. I think that the authors should change this to ``an explicitly spatial analysis of eDNA-derived community similarity''.
\begin{response}
  Indeed, our wording did not clearly convey our meaning. Our intent was to indicate the novelty of the experimental design which was intended to capture relatively fine-scale spatial distribution of eDNA in marine waters. We have made the suggested change (L110).\\
\end{response}

@@

L117 to 122. This whole paragraph can be deleted. This methodology and the steps involved are now commonplace among aquatic biodiversity researchers and need not be enumerated.
\begin{response}
  We appreciate this perspective, but disagree that the method is commonplace to the extent that it can be omitted. Our perspective is that few ecologists understand eDNA methodology, and will likely skip the details we outline in the methods.
	We include this very brief overview for those readers in order to give perspective to the complicated methodology.\\
\end{response}

L166 to 170. Why not use ``µl'' instead of ``microliter''?
\begin{response}
  Our experience is that such characters are often mangled at some point the publication process. The journal is welcome to make this change if they so desire.\\
\end{response}

L169. I guess the authors mean ``1.25 microliter genomic or eDNA template''.
\begin{response}
  We intended `genomic eDNA' to mean the full DNA extracted from samples, but your suggestion is a good one and we have implemented it (L182).\\
\end{response}

L171 to 172. Use ``95 ºC'' instead of ``95C''.
\begin{response}
  We made this change (L154, L159, L162, L184, L185).\\
\end{response}

L195. ``(machine NS500615, run 115, flowcell H3LFLAFXX)''. I don't think we need to know these details!
\begin{response}
  While this does seem like overkill, it is not unheard of for machines or sequencing facilities to develop problems that can bias data, and we included this with the intent of being as transparent as possible. Nevertheless, we removed this information (L209).\\
\end{response}

L196. ``to enhance sequencing depth''. PhiX is added ``to increase sequence diversity'' and not to enhance sequencing depth. Actually, the more PhiX is spiked-in, the less amplicon reads will be obtained, thus decreasing (not enhancing) sequencing depth.
\begin{response}
  We are grateful that you have read the manuscript with such a careful eye!
	Both statements are correct for subtle reasons, and we have incorporated your phrasing to our explanation.
	Because PhiX increases sequence diversity, its effect is to increase the number of clusters which can be correctly differentiated from one another on the flowcell, thus increasing sequencing depth.
	Similar to your statement, the \textbf{less} PhiX is spiked-in, the fewer clusters will be correctly differentiated from one another, and the fewer reads passing filter will be obtained, thus also decreasing sequencing depth (L210).\\
\end{response}

L223. Delete ``a'' in ``This allows for a samples''.
\begin{response}
  We made this change (L237).\\
\end{response}

L270-272. I think that these sentences may be confusing due to the use of some terms related to ``dispersal range'', ``type of fertilization'' and ``type of larval development''. Some ``internally fertilized species'' may have larvae with planktonic development and long dispersal capacity (e.g. some barnacles), whereas some broadcast spawners (e.g. some gorgonians) may be surface brooders with low dispersal ability. I think that the main traits affecting ``dispersal range'' is the type of larval development (planktonic or benthic) and the total planktonic larval duration (PLD), and not the type of fertilization or the strategy for gamete releasing, per se. I think that ``internally fertilized species'' in L271 could be changed to ``benthic brooding species'' and ``broadcast spawners'' could be changed to ``species with long planktonic larval development''. Anyway, this semi-quantitative life-history information recorded by the authors in their data tables does not seem to have been used in their further analyses (since only qualitative considerations regarding two brooding species are discussed below).
\begin{response}
  You make an excellent point that we were not clear in our wording.
	We did not mean to suggest that we made only one general estimate of dispersal per taxon; rather, we made separate estimates for each stage.
	That is, even if a species (e.g. Balanus glandula) is internally fertilized and has sessile adults in the intertidal, if it has a larval stage with high potential for dispersal, we cannot rule out an expectation that it will be most abundant in the sequence data from samples far from shore due to larval abundance.
	We have attempted to clarify this in the text (L284).
	That said, you are also correct that this data is almost unused except to single out Cymatogaster aggregata and Cupolaconcha meroclista as taxa for which an a priori expectation based on adult distribution should be met.
	We added a concluding sentence to the paragraph to explain the reason for these data (L292).
	We chose to keep the description here to show that we have been diligent in investigating potential factors obscuring taxon-specific patterns.
	\\
\end{response}

@@

L285. ``333,537.9 ± 112,200.5''. It looks somehow awkward to use decimal values for numbers of reads here. You could round the values to ``333,538 ± 112,200''
\begin{response}
  Because these are summary statistics (mean and standard deviation), we believe the use of decimals is appropriate.\\
\end{response}

L314-315. You can delete ``the probability that two reads chosen at random from a sample belong to different species'', since it has been already explained in the Materials and Methods section.
\begin{response}
  We made this change (L332).\\
\end{response}

L344. ``For the first time in a marine environment, we document four key patterns''. The authors should definitely use a more modest sentence here. At least the first two patterns they claim to report ``for the first time'' have been already reported previously in several works in the marine environment. E.g.: Fonseca et al. 2014, Guardiola et al. 2015, Lallias et a. 2015, De Vargas et al. 2015, etc., although possibly never with a sampling design specifically designed to get a quantitative assessment of distance decay.
\begin{response}
  Our intent was precisely to convey the novelty of the sampling design; we failed to do that clearly.
	We removed `For the first time in a marine environment' (L264), and we have clarified the novelty of our study in the context of recently published work in the introduction (L76).\\
\end{response}

%------------------------------------REVIEWER2----------------------------------
\subsection*{Reviewer 2 (Stephen Moss)}
\subsubsection*{Basic reporting}
There were a couple of minor grammatical errors in the manuscript. 1) The final sentence (L292-294) of the ``Sequence Data Processing (Bioinformatics)'' paragraph on page 12 in the Results section.
\begin{response}
  We corrected this (L309).\\
\end{response}

2) The first sentence (L359) of the ``Communities far from one another tend to be less similar than those that are nearby'' paragraph on page 15 of the Discussion section.
\begin{response}
  It is unclear what the specific error is, but we acknowledge this sentence is poorly worded.
	We changed this to read `We demonstrate that distant locations have less-similar eDNA communities than proximate locations in Puget Sound, a dynamic marine environment' (L371).\\
\end{response}

I was a bit confused initially over discussion of massively parallel sequencing of 16S amplicons in relation to metazoan diversity in the abstract, and it wasn’t until into the introduction (L98) that I realised it was referring to 16S sequences from the mtDNA. Perhaps some clarity on this earlier on would be helpful?
\begin{response}
  This is a good suggestion; we have modified the abstract to clarify this as early as possible (L9).\\
\end{response}

Perhaps also discussing the reasoning behind the decision to use 16S rDNA from the mtDNA rather than 18S rDNA would be helpful (e.g. Tang et. al., 2012; http://www.pnas.org/content/109/40/16208)?
\begin{response}
  We added a brief explanation of our marker choice to the section Methods:PCR Amplification (L166), and a more general discussion of marker choice to the discussion (L452).\\
\end{response}

@@

I couldn’t find any reference to the details of the assigned clusters in Figure 3. I thought it might be useful to include a reference to figures or tables, perhaps as supplementary information, outlining the taxonomic composition of the clusters?
\begin{response}
	We had a similar first reaction to the results of the classification analysis: \textit{Do these correspond to known ecological communities?}
  The utility of clustering methods is that they simplify the complexity of highly dimensional data; in our case, hundreds to thousands of OTUs, depending on the data subset.
	While additional analyses, such as PCA on a reduced-dimensionality dataset, may help identify which OTUs driving community differentiation, this result will not necessarily correspond to the clusters reported using PAM, especially considering the uncertainty regarding the appropriate number of clusters in our data.
	In short, we are unaware of an analytical or visual method for further simplification of the classification results.
	If you feel this is imperative information, we could potentially modify our \href{https://github.com/jimmyodonnell/Carkeek_eDNA_grid/blob/master/Figures/stacked_bar_community.pdf}{stacked bar plot} in order to indicate the cluster to which each sample was assigned, though making sense of this result in a spatial context is not straightforward.
	\\
\end{response}

I felt the label for Figure 5 needed expanding to provide more detailed information on what it represents, making it more meaningful/understandable as a stand-alone figure.
\begin{response}
  Thank you for noticing this; we have corrected the legend to be meaningful as a standalone figure (Figure 5 legend).\\
\end{response}

The raw data is listed as available under Data Availability, though I assume this is still under embargo? The link on page 18 (L463) will need updating upon publication.
\begin{response}
  Yes; these links were a placeholder until the manuscript is accepted and have now been updated (L514).\\
\end{response}

@@

A GitHub DOI can be provided by Zenodo (https://zenodo.org) and will need updating upon publication.
\begin{response}
  We are aware of this and will create a GitHub version and DOI of the respository once the final version is obtained.\\
	% TODO: upon acceptance, get DOI and update
\end{response}

\subsubsection*{Experimental design}
When detailing how samples were collected (L126-128) under the Environmental Sampling header in the Methods section, I felt that information on how sampling artefacts were mediated (e.g. from L409-414) could have been provided sooner.
\begin{response}
  Perhaps we misunderstand your comment, but in the relevant part of the methods section, we state: `Each environmental sample was collected in a clean 1 liter high-density polyethylene bottle, the opening of which was covered with 500 micrometer nylon mesh to prevent entry of larger particles' (L140).\\
\end{response}

I felt a little more effort could have been made to facilitate the ability to replicate the analyses in this study. The methods section provided good information, but the analyses as detailed on L473 needs more clarity. An outline of the workflow steps and the input/output at each stage would be very helpful here.
\begin{response}
  In our initial draft, we provided an outline of the bioinformatics procedures in the section Methods:Sequence Data Processing (Bioinformatics) (L213), and document the procedures in painstaking detail in the Supplemental Material section Methods:Bioinformatics (L725).
	However, we overlooked the inclusion of a direct link to the scripts used to process the raw sequence data, and have now added those (L515).
	We certainly appreciate any encouragement to move towards more reproducible research, though we are not yet where we would like to be.
	Any specific recommendations are welcome.
	\\
\end{response}

\subsubsection*{Validity of the findings}
In response to the ``(3) Richness declines and evenness increases with distance from shore'' heading in the Discussion section, I felt that decline in richness and increase in evenness might be expected due to dilution and osmotic factors in the ocean. Perhaps some reference to tidal flushing and dilution might be pertinent here?
\begin{response}
  This is an excellent point. We have modified the relevant section to incorporate this potential mechanism (L411).
	\\
\end{response}

I felt that under the remit of the Ecological Analyses headings throughout the study, there could have been some focus on reporting the alpha diversity and rarefaction analysis in relation to whether effective sequencing depth had been achieved for the samples in question.
\begin{response}
  We believe we adequately addressed alpha diversity by reporting richness and evenness both in a 2-dimensional and linear spatial context, with accompanying figures.
	However, we have now also added an OTU accumulation figure to the supplement (Figure 10) and commented on it in the results (L310).
	We note that our goal is not to recover an absolute measure of alpha diversity, but to make comparisons among sites for standardized sampling effort (i.e. sequencing depth).
	\\
\end{response}

### 2017-01-13 Manuscript/v1.1_PeerJ/carkeek_grid.tex
add line numbers to response to reviews

\usepackage[T1]{fontenc} % Choose an output font encoding (T1) that has support for the accented characters used by the most widespread European languages
\usepackage[utf8]{inputenc} % Allow input of accented characters (and more...)
\usepackage{graphicx} %  figures
\graphicspath{ {../../Figures/} {../Figures/} } % include images in the directory 'Figures' at the same level
\usepackage[round]{natbib} % names in citations
\usepackage{lineno} % line numbers
\usepackage{authblk} % allows more intuitive formatting for multiple authors/affiliations
\usepackage[margin=1in]{geometry} % make the margins 1 inch on all sides of the document
%\usepackage{amsmath} % useful for formatting math stuff, especially complex equations
%\usepackage{pdflscape} % rotate table into landscape mode
%\usepackage{subfigure} % side by side figures
%\usepackage{longtable} % for tables that span multiple pages.
\usepackage{setspace} % double spacing
\usepackage{float} % figure placement [H]
\usepackage{nameref,hyperref} % named references (e.g. "Supplemental Material")

@@

\subsection*{PCR Amplification}
We chose a primer set that amplifies an approximately 115 base pair (bp) region of the mitochondrial 16S rRNA gene in at least 10 metazoan phyla from this habitat, excludes non-metazoans, and resolves taxonomy to the family level in most cases using a public sequence database \citep{Kelly2016FMS}.
We used a two-step polymerase chain reaction (PCR) protocol described by \citet{ODonnell2016} to generate 4 replicate products from each DNA sample.
In the first set of reactions, primers were identical in every reaction (forward: AGTTACYYTAGGGATAACAGCG; reverse: CCGGTCTGAACTCAGATCAYGT); primers in the second set of reactions included these same sequences but with 3 variable nucleotides (NNN) and an index sequence on the 5$'$ end (see \nameref{sequencing_metadata}). We used the program OligoTag \citep{Coissac2012} to generate 30 unique 6-nucleotide index sequences differing by a minimum Hamming distance of 3 (see \nameref{sequencing_metadata}). Indexed primers were assigned to samples randomly, with the identical index sequence on the forward and reverse primer to avoid errors associated with dual-indexed multiplexing \citep{Schnell2015}. In a UV-sterilized hood, we prepared 25 microliter reactions containing 18.375 microliters ultrapure water, 2.5 microliters 10x buffer, 0.625 microliters deoxynucleotide solution (8 millimolar), 1 microliter each forward and reverse primer (10 micromolar, obtained lyophilized from Integrated DNA Technologies (Coralville, IA, USA)), 0.25 microliters Qiagen HotStar Taq polymerase, and 1.25 microliter genomic eDNA template at 1:100 dilution in ultrapure water. PCR thermal profiles began with an initialization step (95 ºC; 15 min) followed by cycles (40 and 20 for the first and second reaction, respectively) of denaturation (95 ºC; 15 sec), annealing (61 ºC; 30 sec), and extension (72 ºC; 30 sec). 20 identical PCRs were conducted from each DNA extract using non-indexed primers; these were pooled into 4 groups of 5 in order to ensure ample template for the subsequent PCR with indexed primers. In order to isolate the fragment of interest from primer dimer and other spurious fragments generated in the first PCR, we used the AxyPrep Mag FragmentSelect-I kit with solid-phase reversible immobilization (SPRI) paramagnetic beads at 2.5x the volume of PCR product (Axygen BioSciences, Corning, NY, USA). A 1:5 dilution in ultrapure water of the product was used as template for the second reaction. PCR products of the second reaction were purified using the Qiagen MinElute PCR Purification Kit (Qiagen, Hilden, Germany). Ultrapure water was used in place of template DNA and run along with each batch of PCRs to serve as a negative control for PCR; none of these produced visible bands on an agarose gel. In total, four separate replicates from each of 31 DNA samples were carried through the two-step PCR process for a total of 124 sequenced PCR products. These were combined with additional samples from other projects, totaling 345 samples for sequencing.

>>>>

\usepackage[T1]{fontenc} % Choose an output font encoding (T1) that has support for the accented characters used by the most widespread European languages
\usepackage[utf8]{inputenc} % Allow input of accented characters (and more...)
\usepackage{graphicx} %  figures
\graphicspath{ {../../Figures/} {../Figures/} } % include images in the directory 'Figures' at the same level
\usepackage[round]{natbib} % names in citations
\usepackage{lineno} % line numbers
\usepackage{authblk} % allows more intuitive formatting for multiple authors/affiliations
\usepackage[margin=1in]{geometry} % make the margins 1 inch on all sides of the document
%\usepackage{amsmath} % useful for formatting math stuff, especially complex equations
%\usepackage{pdflscape} % rotate table into landscape mode
%\usepackage{subfigure} % side by side figures
%\usepackage{longtable} % for tables that span multiple pages.
\usepackage{setspace} % double spacing
\usepackage{float} % figure placement [H]
\usepackage{nameref} % named references (e.g. "Supplemental Material")
\usepackage{breakcites} % add line breaks in citations where necessary
\usepackage[breaklinks=true]{hyperref}

@@

\subsection*{PCR Amplification}
We chose a primer set that amplifies an approximately 115 base pair (bp) region of the mitochondrial 16S rRNA gene in at least 10 metazoan phyla from this habitat, excludes non-metazoans, and resolves taxonomy to the family level in most cases using a public sequence database \citep{Kelly2016FMS}.
We used a two-step polymerase chain reaction (PCR) protocol described by \citet{ODonnell2016} to generate 4 replicate products from each DNA sample.
In the first set of reactions, primers were identical in every reaction (forward: AGTTACYYTAGGGATAACAGCG; reverse: CCGGTCTGAACTCAGATCAYGT); primers in the second set of reactions included these same sequences but with 3 variable nucleotides (NNN) and an index sequence on the 5$'$ end (see \nameref{sequencing_metadata}). We used the program OligoTag \citep{Coissac2012} to generate 30 unique 6-nucleotide index sequences differing by a minimum Hamming distance of 3 (see \nameref{sequencing_metadata}). Indexed primers were assigned to samples randomly, with the identical index sequence on the forward and reverse primer to avoid errors associated with dual-indexed multiplexing \citep{Schnell2015}. In a UV-sterilized hood, we prepared 25 microliter reactions containing 18.375 microliters ultrapure water, 2.5 microliters 10x buffer, 0.625 microliters deoxynucleotide solution (8 millimolar), 1 microliter each forward and reverse primer (10 micromolar, obtained lyophilized from Integrated DNA Technologies (Coralville, IA, USA)), 0.25 microliters Qiagen HotStar Taq polymerase, and 1.25 microliter genomic or eDNA template at 1:100 dilution in ultrapure water. PCR thermal profiles began with an initialization step (95 ºC; 15 min) followed by cycles (40 and 20 for the first and second reaction, respectively) of denaturation (95 ºC; 15 sec), annealing (61 ºC; 30 sec), and extension (72 ºC; 30 sec). 20 identical PCRs were conducted from each DNA extract using non-indexed primers; these were pooled into 4 groups of 5 in order to ensure ample template for the subsequent PCR with indexed primers. In order to isolate the fragment of interest from primer dimer and other spurious fragments generated in the first PCR, we used the AxyPrep Mag FragmentSelect-I kit with solid-phase reversible immobilization (SPRI) paramagnetic beads at 2.5x the volume of PCR product (Axygen BioSciences, Corning, NY, USA). A 1:5 dilution in ultrapure water of the product was used as template for the second reaction. PCR products of the second reaction were purified using the Qiagen MinElute PCR Purification Kit (Qiagen, Hilden, Germany). Ultrapure water was used in place of template DNA and run along with each batch of PCRs to serve as a negative control for PCR; none of these produced visible bands on an agarose gel. In total, four separate replicates from each of 31 DNA samples were carried through the two-step PCR process for a total of 124 sequenced PCR products. These were combined with additional samples from other projects, totaling 345 samples for sequencing.
